Optical characterization of gaps in directly bonded Si compound optics using infrared spectroscopy

### 2015-01-23 AO_bondedSi.tex

In the limit $\lim_{d \to 0} T_e \rightarrow T_{DSP}} $ and the etalon approaches 100\% transmission.  Figure \ref{figAbsoluteTrans} shows a plot of Equation \ref{eqFP} for gap sizes, $d$, of 0, 20, 35, 60, 350, and 3500 nm.  The bottom two panels show the ``gap'' transmission, the value of the throughput with the losses from the outer Si cavities divided out.  For the very largest gap, large fringes appear as the cavity resonance of the air gap modulates the transmission.  Below a $\lambda/2$ gap size the effect of the gap manifests itself as a wavelength-dependent diminution of the transmission.  For the smallest $35-60$ nm gaps, there is a $2-6\%$ decrease in gap transmission from 2400 nm down to 1250 nm.  It is the magnitude and wavelength dependence of this transmission spectrum that provides information about the axial extent of the gap.

\section{Laboratory demonstration: Imbedding gaps of known sizes}

We evaluated our metrology method by imbedding gaps of known sizes in the interface.  Here we describe the process of how we made the gaps and how we characterized their axial extent.

The substrate thicknesses ranged from 0.5 to 3.3 mm per substrate.  We classify substrates with thicknesses larger than 1 mm as ``pucks''.  Substrates with thickness less than 1.0 mm thick are called ``wafers''.  The distinction between the two classes of substrate thickness is based on the anticipation that thin substrates will conform more easily to their bonding partner substrates.  There is also evidence that bond front propagation is slower in thick pucks \cite{2007ApOpt..46.6793H} than in wafers.  Our ultimate goal is to bond optics with thicknesses as large as 30 mm.  These pucks truly resemble over-sized hockey pucks.

The three important characteristics of the gaps are their axial extent, their lateral areal extent, and their areal fill factor.  We bored holes in silicon with inductively coupled plasma etching.  We used two different gases with low and high etch rates.  CHF$_3$ on Si exhibited about 0.3 nm/second etch rate, SF$_6$ exhibited about 1 micron/minute.  The area exposed to the process gases was either a 25 mm diameter circle, or a 15 mm $\times$ 20 mm rectangle.  We protected the unexposed areas with thick photoresist.  We achieved etch depths of 15$\pm$5 nm and 4000$\pm$100 nm, as measured with Veeco NT9100 Optical Profiler for the small depth, and Dektak stylus profilometry for the large depth.  Table \ref{tabbondexper} has more information about the bonding samples, and Table \ref{tabPatternFills} lists the properties of the patterns.

The fill factor is defined as the pattern area covered by gaps relative to the total pattern area.  In a second round of experiments, we achieved gap sizes of 14$-$95 nm on four substrates with three meshes with coarse, medium, and fine boxes, each with 50\% fill factors.  No attempt was made to verify the delivered fill factor, but since high precision lithographic processes were employed, we assign no uncertainty to the difference between the delivered fill factor and the designed fill factor.  For measurements taken with the Veeco NT9100 Optical Profiler, uncertainties were constructed by inspecting the histogram of topology, as shown in Figure \ref{figVS20pattern}.  The large-scale distortions were removed from the topology by masking and flattening to ensure that the uncertainties accurately reflect the distribution of measured heights.  The stylus profiliometry measured values were simply assigned an uncertainty of 5\%, which was consistent with our intuition and experience.  Table \ref{tabbondexper} lists part numbers and descriptions for our bonded substrates.

\begin{figure}[htbp]
\centerline{\includegraphics[width=1.0\columnwidth]{figs/VS20fineGapCrop.pdf}}
\caption{
\label{figVS20pattern}
Veeco optical profiler surface plot and histogram for part VS20 in the fine pattern section.  The 100 mm diameter VS20 is comprised of two thick silicon wafers.  We patterned three meshes with coarse, medium, and fine boxes, each with 50\% fill factors.  The depth was 95 $\pm$ 5 nm as indicated by the well-separated peaks in the Veeco profilometry histogram.  The fill factor for this and all mesh patterns was 50\%.  The purpose for these patterns was to have gaps of known axial extent against which we could test the optical metrology technique we describe in this article.}
\end{figure}

@@

The Cary 5000 has four operation modes: single front, single back, double, and double reverse.  We experimented with single front and double modes.  Double mode takes a reference spectrum simultaneously with the target spectrum.  In both single and double mode, a baseline scan is taken with no sample in the measurement holder.  The baseline spectrum is the raw spectral response of the system.  We automatically divide sample spectra by the baseline spectrum.  In single beam mode, the baseline measurements must be repeated about every half hour for post-process division.  In double beam mode the front beam is automatically divided by the rear beam, which corrects out lamp luminosity drift, for example.

@@

The spectral sampling was 2.0 nm.  We used a spectral resolution of 5.0 nm, the lowest spectral resolution the Cary5000 offers.  We expect smoothly varying spectral features for all conceivable interface gaps in Si bonded wafers.  Specifically, Fabry-P\`erot fringes attributable to interfacial gaps in bonded Si will be well-sampled with 5 nm spectral resolution, so long as the interface gap is less than $\sim$100 $\mu$m.  The axial extends of such large gaps would be easily measurable by other means, like IR imaging.

@@

We looked for the indication of large gaps detectable as ``IR bubbles'' \cite{1992JEMat..21..669M} with IR imaging.  The detector was an IR Vista $\alpha$NIR infrared focal plane array, with 312 $\times$ 252 pixels, with 30 $\mu$m square pixels.  We used a 50 mm F/2 lens, with a fiber optic white light illumination source.  The combination of the absorption of Si, the lamp spectrum and the detector responsivity led to an effective bandpass from approximately 1.15$-$1.6 $\mu$m.  The field of view was smaller than the 100 mm diameter wafers, so we dithered the sample.  Figure \ref{figVS2021_IR_image} shows an IR image of sample VS20-21.  For this image we coarsely flat fielded the detector by observing a homogeneously illuminated white screen. Despite the flat fielding, detector non-uniformities and vignetting are perceptible in our image as vertical stripes.  Mosaic-stitching errors are also perceptible.  It was easy to detect IR bubbles despite detector artifacts since the bubbles translate with the sample when we dither its position.  We detected IR bubbles in all 3 of the bonded Si samples on which we recorded IR images.  The bubble areal density varied from about 4 per 100 mm diameter bonded wafer pair to over 20 per bonded wafer pair.  The individual IR bubble diameters were as small as a few square millimeters up to 400 mm$^2$.  The morphologies of the IR bubbles ranged from perfectly circular to irregular.  The structure within the IR bubbles was sometimes indicative of Newton's rings, exposing the presence of gaps larger than about $\lambda/2$ up to $\sim 15 \lambda$.  As expected, the unbonded DSP wafers show no such IR bubbles.  No attempt at quantitative measurement was made on the IR images due to evidence of a non-linear detector response.  The bulk morphologies were clear regardless of the detector response.  When IR imaging was available, it was leveraged to target IR specectrscopy in locations demonstrating presence or absense of IR bubbles, as desired.  In other cases when IR imaging was not available or ignored, spectra were taken at random positions on the wafer.

\begin{figure}[htbp]
\centerline{\includegraphics[width=0.4\columnwidth]{figs/VS2021_IR_image.pdf}}
\caption{
\label{figVS2021_IR_image}
IR image of part VS20-21. Three rectangular mesh areas are apparent (Table \ref{tabPatternFills}).  The coarsest mesh, C, shows substructure, while the medium and fine meshes are indistinguishable at the low resolution of the image.  The three dark spots are gaps at the inferface of the direct bonded Si samples.}
\end{figure}

\section{Results of infrared spectroscopy of directly bonded Si}
\label{secResults}
In this section we describe the spectroscopy of directly bonded Si.

@@

\subsection{Inferring axial extent of gaps from noisy transmission spectra}
We infer the axial extent of gaps from their noisy transmission spectra.  We use a Bayesian inference technique.  First we define a mixture model in which the observed normalized gap spectrum $T_{obs}$ is composed of a sum of a spectrum through a gap of axial extent $d$, and a spectrum through a perfect bond:

@@

\begin{figure}[htbp]
        \centering
        \begin{subfigure}[b]{0.5\textwidth}
              \includegraphics[width=\textwidth]{figs/VG12_corner.pdf}
              \caption{Fitted gap size $d$ and fill factor $f$ for VG12}
		\label{figVG12_corner}
        \end{subfigure}

        \begin{subfigure}[b]{0.5\textwidth}
                \includegraphics[width=\textwidth]{figs/VG12_f50.pdf}
                \caption{Measured spectrum and fit of VG12 position 2}
                \label{figVG12_f50}
        \end{subfigure}
\caption{ Same as Figure \ref{figVG03full} and \ref{figVG03part}, but for sample VG12, which has an imbedded gap of axial extent $d=49\pm6\;$nm with fill factor $f=0.5$.  The measured spectrum in this region shows a clear deficit of transmission from the gapless counterparts.  The measured spectrum also shows a deficit of transmission compared to the predicted transmission of a 49 $\pm6\;$nm gap with 50\% fill factor (green apple shaded color band).  The discord between prediction and measurement could be due to several factors including an underestimate of the intrinsic gap size, an underestimate of the intrinsic fill factor, or an additional gap in excess of the prediction.  We favor the latter interpretation, since undesired gaps are common in these direct-bonded Si samples that have not undergone high-temperature thermal annealing.
\label{figVG12}}
\end{figure}

>>>>

In the limit $\lim_{d \to 0} T_e \rightarrow T_{DSP}} $ and the etalon approaches 100\% transmission.  Figure \ref{figAbsoluteTrans} shows a plot of Equation \ref{eqFP} for gap sizes, $d$, of 0, 20, 35, 60, 350, and 3500 nm.  The bottom two panels show the ``gap'' transmission, the value of the throughput with the losses from the outer Si cavities divided out.  For the very largest gap, large fringes appear as the cavity resonance of the air gap modulates the transmission.  Below a $\lambda/2$ gap size the effect of the gap manifests itself as a wavelength-dependent diminution of the transmission.  For the smallest $35-60$ nm gaps, there is a $2-6\%$ decrease in gap transmission from $2400\;$nm down to $1250\;$nm.  It is the magnitude and wavelength dependence of this transmission spectrum that provides information about the axial extent of the gap.

\section{Laboratory demonstration: Embedding gaps of known sizes}

We evaluated our metrology method by embedding gaps of known sizes in the interface and measuring the effects of these gaps on IR transmission.  The substrate thicknesses ranged from 0.5 to 3.3 mm.  We denote substrates with thicknesses larger than 1 mm as ``pucks'' and substrates with thickness less than 1.0 mm thick as ``wafers''.  We distinguish the two classes of substrate thickness based on the anticipation that thin substrates will conform more easily to their bonding partner substrates.  There is also evidence that bond front propagation is slower in thick pucks \cite{2007ApOpt..46.6793H} than in wafers.  Our ultimate goal is to bond optics with thicknesses as large as 30 mm.

The three important characteristics of gaps are their axial extent, their lateral areal extent, and their areal fill factor.  To fabricate our test samples, we used photoresist lithography to pattern Si substrates, and we bored holes with inductively coupled plasma etching.  We used two different gases with low and high etch rates, CHF$_3$ on Si exhibited about 0.3 nm/second etch rate, SF$_6$ exhibited about 1 micron/minute.  Tables \ref{tabbondexper} and \ref{tblPatternFills} give information about the hole patterns and depths, as measured with Veeco NT9100 Optical Profiler for the small depth, and Dektak stylus profilometry for the large depth.

The fill factor is defined as the pattern area covered by gaps relative to the total pattern area.  In a second round of experiments, we achieved gap sizes of 14$-$95 nm on four substrates with three meshes with coarse, medium, and fine boxes, each with 50\% fill factors.  No attempt was made to verify the delivered fill factor, but since high precision lithographic processes were employed, we assign no uncertainty to the difference between the delivered fill factor and the designed fill factor.  For measurements taken with the Veeco NT9100 Optical Profiler, uncertainties were constructed by inspecting the histogram of topology, as shown in Figure \ref{figVS20pattern}.  The large-scale distortions were removed from the topology by masking and flattening to ensure that the uncertainties accurately reflect the distribution of measured heights.  The stylus profilometry measured values were simply assigned an uncertainty of 5\%, which was consistent with our intuition and experience.  Table \ref{tabbondexper} lists part numbers and descriptions for our bonded substrates.

\begin{figure}[htbp]
\centerline{\includegraphics[width=1.0\columnwidth]{figs/VS20fineGapCrop.pdf}}
\caption{
\label{figVS20pattern}
Veeco optical profiler surface plot and histogram for part VS20-21 in the fine pattern section.  The 100 mm diameter VS20-21 is comprised of two thick silicon wafers.  We patterned three meshes with coarse, medium, and fine boxes, each with 50\% fill factors.  The depth was 95 $\pm$ 5 nm as indicated by the well-separated peaks in the Veeco profilometry histogram.  The fill factor for this and all mesh patterns was 50\%.  The purpose for these patterns was to have gaps of known axial extent against which we could test the optical metrology technique we describe in this article.}
\end{figure}

@@

The Cary 5000 has four operation modes: single front, single back, double, and double reverse.  We experimented with single front and double modes.  Double mode takes a reference spectrum simultaneously with the target spectrum.  In both single and double mode, a baseline scan is taken with no sample in the measurement holder.  We automatically divide sample spectra by the baseline spectrum.  In single beam mode, the baseline measurements must be repeated about every half hour for post-process division.  In double beam mode the front beam is automatically divided by the rear beam, which corrects out lamp luminosity drift, for example.

@@

The spectral sampling was ranged from 2.0 to 10.0 nm.  We used a spectral resolution of 5.0 nm.  We expect smoothly varying spectral features for all conceivable interface gaps in Si bonded wafers.  Specifically, Fabry-P\`erot fringes attributable to interfacial gaps in bonded Si will be well-sampled with 5 nm spectral resolution, so long as the interface gap is less than $\sim$100 $\mu$m.  The axial extents of such large gaps would be easily measurable by other means, like IR imaging.

@@

We looked for large gaps, detectable as ``IR bubbles'' \cite{1992JEMat..21..669M} with IR imaging.  The detector was an IR Vista $\alpha$NIR infrared focal plane array, with 312 $\times$ 252 pixels, with 30 $\mu$m square pixels.  We used a 50 mm F/2 lens, with a fiber optic white light illumination source.  The combination of the absorption of Si, the lamp spectrum and the detector responsivity led to an effective bandpass from approximately 1.15$-$1.6 $\mu$m.  The field of view was smaller than the 100 mm diameter wafers, so we dithered the sample.  Figure \ref{figVS2021_IR_image} shows an IR image of bonded sample VS20-21.  For this image, we coarsely flat-fielded the detector by observing a homogeneously illuminated white screen. Despite the flat-fielding, detector non-uniformities and vignetting are perceptible in our image as vertical stripes.  Mosaic-stitching errors are also perceptible.  It was easy to detect IR bubbles despite detector artifacts since the bubbles translate with the sample when we dither its position.

We also used IR imaging to inspect VG09-12 and VG15-13.  We detected IR bubbles in all 3 of these bonded Si samples.  The observable bubble areal density varied from about 4 per 100 mm diameter bonded wafer pair to over 20 per bonded wafer pair.  The individual IR bubble diameters were as small as a few square millimeters up to 400 mm$^2$.  The IR bubbles had varied morphologies.  The structure within the IR bubbles was sometimes indicative of Newton's rings, exposing the presence of gaps larger than about $\lambda/2$ up to $\sim 15 \lambda$.  As expected, the unbonded DSP wafers show no such IR bubbles.  No attempt at quantitative measurement was made on the IR images due to evidence of a non-linear detector response.  The bulk morphologies were clear regardless of the detector response.  When IR imaging was available, it was leveraged to target IR specectroscopy in locations demonstrating presence or absense of IR bubbles, as desired.  In other cases when IR imaging was not available or ignored, spectra were taken at random positions on the wafer.

\begin{figure}[htbp]
\centerline{\includegraphics[width=0.4\columnwidth]{figs/VS2021_IR_image.pdf}}
\caption{
\label{figVS2021_IR_image}
IR image of part VS20-21. Three rectangular mesh areas are apparent (Table \ref{tabPatternFills}).  The coarsest mesh, C, shows substructure, while the medium and fine meshes are indistinguishable at the low resolution of the image.  The three dark spots are gaps at the interface of the direct bonded Si samples.}
\end{figure}

\section{Results of infrared spectroscopy of directly bonded Si}
\label{secResults}

@@

The gaps in our test pairs should be a combination of the deliberately induced gaps and any inadvertent additional spaces caused by imperfections in the bonds.  We infer the axial extent of gaps in our samples with etched holes from their measured transmission spectra.  We use a Bayesian inference technique.  First we define a mixture model in which the observed normalized gap spectrum $T_{obs}$ is composed of a sum of a spectrum through a gap of axial extent $d$, and a spectrum through a perfect bond:

@@

\begin{figure}[htbp]
        \centering
        \begin{subfigure}[b]{0.5\textwidth}
              \includegraphics[width=\textwidth]{figs/VG12_corner.pdf}
              \caption{Fitted gap size $d$ and fill factor $f$ for VG09-12}
		\label{figVG12_corner}
        \end{subfigure}

        \begin{subfigure}[b]{0.5\textwidth}
                \includegraphics[width=\textwidth]{figs/VG12_f50.pdf}
                \caption{Measured spectrum and fit of VG09-12 position 2}
                \label{figVG12_f50}
        \end{subfigure}
\caption{ Same as Figure \ref{figVG03full} and \ref{figVG03part}, but for sample VG09-12, which has an imbedded gap of axial extent $d=49\pm6\;$nm with fill factor $f=0.5$.  The measured spectrum in this region shows a clear deficit of transmission from the gapless counterparts.  The measured spectrum also shows a deficit of transmission compared to the predicted transmission of a 49 $\pm6\;$nm gap with 50\% fill factor (green apple shaded color band).  The discord between prediction and measurement could be due to several factors including an underestimate of the intrinsic gap size, an underestimate of the intrinsic fill factor, or an additional gap in excess of the prediction.  We favor the latter interpretation, since undesired gaps are common in these direct-bonded Si samples that have not undergone high-temperature thermal annealing.
\label{figVG12}}
\end{figure}
